---
title: "Weekly assignment 1"
author: "Matt Ring"
date: "2021-08-28"
output: html_document
---

This is an R Markdown file. Below there are several things that I will ask you to add or modify. Along the way, I'm going to include new R Markdown methods to help expand your R Markdown toolkit.

Please start by editing the YAML section above. Add your own name and the date you will submit this assignment.

Please save this document in your project folder, inside of `output/markdown`.

**!Important**: Before knitting this file, we'll have to add a couple of libraries to your project packages (packrat). Please run the following in your console pane:

```{r install_packages, eval = FALSE}
install.packages(
  c('kableExtra',
    'DT', 
    'tidyverse'))
```

The above is called a "code chunk".

In the curly brackets (`{}`) above:

* The first argument species the program that you are calling, which is in this case `r`. 
* The second argument, which is separate by a space from your program argument, provides a label for your code chunk.
* The next argument, separated by a comma, provides chunk options. In this example, it's saying not to run the code chunk in R. *Note: If eval were set to TRUE, it would try to install the packages each time the document is knit!!*

*Note: the html tag `<hr>` below creates a horizontal line.*

<hr>

## Code chunks

The code chunk below is the setup section for the code chunks in the document. Code chunks begin with three back-ticks, then curly braces, a set of instructions for the chunk, and options for running the chunk or styling the output. The instructions here say to run Program R, call the chunk "setup", and don't include the code in the knitted document. Knit the document to see how it looks.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The code chunk below loads the tidyverse meta-package. This is necessary, because it allows the document to call on tidyverse functions and data.

Note that the chunk option `message=TRUE` is actually the default behavior. Change this option to FALSE and knit the document to see what happens. **(0.5 points)**

```{r, message=FALSE}
library(tidyverse)
```

**Answer:** When `message = FALSE`, `library(tidyverse)` no longer outputs which packages were loaded and what conflicts occured.

### Add a new code chunk

The data set `fruit` is a character vector that became available in this session if you correctly loaded the tidyverse package.

Insert a new code chunk (Windows: Ctrl+Alt+I, Mac: Command+Option+I). Inside the chunk, use indexing to subset the `fruit` vector to the first four values and assign the name `my_fruit` to the resultant object. **(0.5 points)**

```{r}
my_fruit <- fruit[1:4]
```

### Code chunks for styling css:

We can use code chunks to modify the css of our document (cascading style sheet). Below, I've added some styling to this document using css **but** I've commented out each style element. For each style element knit the document then un-comment that section and knit it again to see how the document changes. **(0.75 points)**

* We can change a single element like this. Here, I change the font size of the html tag p to 18. `<p>` makes a paragraph.

```{css css_p}
/*here is how we add a comment in css*/

/*p {
  font-size: 18px;
}*/
```

**Answer:** The above code chunk makes paragraph text much larger, but not lists or headers.

* You can style multiple elements at once, separating each tag with a comma. `<ul>` is the tag for unordered list (bullet points), and `<ol>` is the tag for ordered lists (numbers).

```{css css_lists}
/*ul, ol {
  font-size: 18px;
}*/
```

**Answer:** Unordered and ordered lists both have visibly larger text when the above code chunk is uncommented.

* `<pre>` is the html tag for code blocks
```{css css_pre}
/*pre {
  margin-top: 30px;
  margin-bottom: 30px;
  background-color: #e6f2ff;
}*/
```
**Answer:** The above code chunk changed the background color of each code block to light blue. The distance between code chunks and other objects in this RMarkdown file increased as well.

Add another css code chunk below. The html tags for items in a list is `<li>`. Provide a unique label (of your choice) to your chunk. In your code chunk, add padding to each list item. Set the padding-top to 8px. **(0.5 points)**

```{css css_student_made}
/*Adds padding above lists*/

li {
  padding-top: 8px
}

```

<hr>

## About you (text elements)

I like to fine tune course material to suit the needs and interests of the students. 

### Your bio

Please delete my information in the list below and add your own. **(1.5 points)**

* **My research focus/interest**: I'm a generalist, but have recently been working on zoning in the DC area and transit policy. Will be switching to labor policy with the Bureau of Labor Statistics soon.
* **I am taking this class because**: I find location data makes for very practical and interesting communicable reports. It's also something I've worked with frequently.
* **The topic I would most like to learn in this class**: Where to find GIS data.
* **One fun fact about me**: I really enjoy gardening, but only got started once moving to DC. I currently have some parsley, basil, oregeno, and chives growing on my back porch.
* **One thing that people like about me**: I'm very bad at lying, to the point where I ALWAYS lose at the card game BS.
* **One thing that people find annoying about me**: I talk about Michigan (my home state) a lot. I also have zero capacity to multi-task, I probably can't "walk and chew gum". 

You can add a comment in an html document using `<!--- my comment --->`. For example, perhaps you don't appreciate being asked why people think you are annoying. You could sneak a comment to me like this:

* **One thing that people find annoying about me**: Trick question, no one is annoyed by anything that I do<!--- I do not appreciate this question Brian --->

See what happens when you knit this!

**Answer:** The text disappears!

### Your computational skillset

It's also important to get a sense of the skills that students are coming in with. Below is a list of R tasks. Please modify this list by coloring a list item red if you have no previous experience with the list item. **(1 point)**

* <span style = "color: red;">**Not** being a super weirdo about R script formatting</span>
* ggPlot
* <span style = "color: red;">Writing in html</span>
* <span style = "color: red;">css</span>
* Shiny apps
* Mutating data with dplyr
* Data normalization
* <span style = "color: red;">Packrat</span>
* Filtering data with dplyr
* Bash commands
* Joining data with dplyr
* Using magrittr's pipe operator
* Making Shiny apps
* The RStudio IDE
* <span style = "color: red;">Using R to analyze raster data</span>
* Leaflet
* <span style = "color: red;">Lubridate</span>
* <span style = "color: red;">Regex</span>
* Working in R Studio projects
* Stringr
* <span style = "color: red;">Creating video lectures for this class</span>

### A picture of you

You can use the html `<img>` tag to create a picture. Here, I'd like a picture of you, your pet, or something you like. Modify the html below to add your picture and provide your own caption. For this document to be share-able as is, you'll have to include a picture that is available on the internet. **(2 points)**

<figure>
  <img 
    src = "https://eoimages.gsfc.nasa.gov/images/imagerecords/88000/88256/sleepingbear_oli_2016169_detail.jpg" 
    style = "display: block;
      margin-top: 24px;
      margin-left: auto;
      margin-right: auto;
      width: 90%;">
    <figcaption><br>A photo related to both my "annoying" trait and maps! My grandparents used to live in this area of Michigan and I have fond memories of exploring Sleeping Bear Dunes.</figcaption>
</figure>

**Please note**: If you have your pictures stored in Google Photos you can use <a href = "https://www.labnol.org/embed/google/photos/" target = "_blank">this webpage</a> to get a embed link to a Google Photos. **Hey! Notice how we just created a hyperlink?**

<figure>
  <img 
    src = "https://lh3.googleusercontent.com/2GUz4c_oU-OgUAQxHswmJ3hBd60rudw7IkeELoIpV6LkJOUQh5moT7EkNpvqD8nf1g-yijbkZnFuKNEC50yjwRyCulgjtuQ5DU19EmkQQmHRpG8x5Vaw6kLU7dBOzxp4Ftq23Zt0cNY=w2400" 
    style = "display: block;
      margin-top: 24px;
      margin-left: auto;
      margin-right: auto;
      width: 90%;">
    <figcaption><br>This is Doris the Dog.</figcaption>
</figure>

<hr>

## Playing in the band (tables)

You can see data sets that are available to you (either in base R or in your packages) by typing `data()` in your console. Data that are in these packages can be called just by typing their name.

We'll look at the `band_instruments` data set, which is in the dplyr package. Insert a new R code chunk and type the name of the data set to print its contents. 

```{r}
# Displays the band instruments dataset
band_instruments
```


What if we wanted to see this as a nice html table though? There are a few options for doing so. Most are awful to deal with. I like using `knitr::kable()`:

```{r basic_kable}
knitr::kable(band_instruments)
```

### Nicer tables with kableExtra

Now let's make a nicer table using the tools in the library kableExtra. We can use `kableExtra::kable_styling()` to modify our table.

```{r better_kable}
band_kable <-
  knitr::kable(band_instruments,
               format = 'html')

kableExtra::kable_styling(band_kable, 'striped')
```

### Even nicer tables with the Data table package

Another option, which is a preference of mine, is to use `DT::datatable`. DT is a package that provides an R interface to the JavaScript DataTables library.

```{r dt}
DT::datatable(band_instruments)
```

It's a little silly in this context. We have a search box and pagination for a table with only three rows. The great thing about the DataTables JavaScript package is that you've got tons of controls. Here, I'll remove the paging and search feature:

```{r better_dt}
DT::datatable(band_instruments,
              options = list(paging = FALSE,
                             searching = FALSE))
```

Use `data(package = 'ggplot2')` to see what data sets are in ggplot. I found the data set `mpg` -- that's a pretty good one for this example. It has fuel economy data for cars for 1999 and 2008.

```{r data_search}
# Shows different data packages available via ggplot2
data(package = 'ggplot2')
```


```{r mpg_tibble}
mpg
```

Add a code chunk that uses `DT::datatable()` to print the `mpg` data frame into an html table. Provide the unique label, of your choice, to the chunk. **(2 points)**

```{r display_mpg}
# Displays the mpg dataset 
DT::datatable(mpg)

```


## Fuel economy (data visualization)

Probably one of my favorite things about generating R Markdown documents is that it's super easy to put your data visualizations in there. Here's a really ugly example of a plot:

```{r ugly_ggplot}
ggplot(
  data = mpg,
  aes(x = class,
      y = hwy)) +
  geom_boxplot()
```

Add a code chunk that creates a scatterplot (replace `geom_boxplot()` with `geom_point()`) to visualize how highway fuel economy (column = `hwy`) is associated with city fuel economy (column = `cty`). You can style you plot however you would like and add any plot elements that you find interesting. **(1.25 points)**

```{r new_ggplot}
# Creates a plot of city fuel economy vs. highway fuel economy
ggplot(
  data = mpg,
  aes(x = cty,
      y = hwy)) +
  geom_point() +
  labs(title = "Highway vs. City Fuel Economy",
       x = "City Fuel Economy (mpg)",
       y = "Highway Fuel Economy (mpg)") +
  # Adds an overall theme to the visualization
  theme_minimal()
```






